---
title: Getting Started with neuroatlas
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts
- figures

vignette: |
  %\VignetteIndexEntry{Getting Started with neuroatlas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "ragg_png")
}
if (requireNamespace("systemfonts", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  albersdown::albers_register_fonts()
}
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(
    albersdown::theme_albers(
      family = params$family,
      preset = params$preset
    )
  )
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 4.5
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```


You have a brain image and want to ask a regional question: *what value does
this image take in each parcel, region, or hemisphere?* The difficult part is
not computing a mean. It is keeping the atlas labels, spatial grid, metadata,
and plotting representation aligned from beginning to end.

This article follows one complete path. You will load the bundled FreeSurfer
ASEG atlas, inspect its regions, create a compatible image, reduce that image to
one value per region, and draw the atlas. The core path is offline and every
result is executed when this vignette is built.

## What are the objects?

Three object transitions organise the workflow:

| You have | What it contains | A common next step |
|---|---|---|
| an `atlas` | a labelled volume plus one row of metadata per region | `roi_metadata()`, `get_roi()`, or `reduce_atlas()` |
| a `NeuroVol` | one numeric value at every voxel on a declared grid | summarise it with a compatible volume atlas |
| a regional tibble | one value per parcel or anatomical region | analyse it or pair it with a matching plot |

A `surfatlas` is the surface counterpart of an `atlas`. It stores a labelled
mesh for each hemisphere rather than a labelled volume. Volume and surface
atlases share region IDs and metadata, but they are not interchangeable
containers.

## How do you load and inspect an atlas?

`get_atlas()` is the general discovery-and-loading entry point. Atlas-specific
loaders remain useful when you want their specialised arguments.

```{r load-atlas}
library(neuroatlas)

atlas <- get_atlas("aseg")
atlas
```

The object contains 17 subcortical and midline regions on an
`MNI152NLin6Asym` grid. Prefer public accessors over reaching into the object
when an accessor exists:

```{r inspect-metadata}
meta <- roi_metadata(atlas)
meta[c("id", "label", "hemi")]
```

```{r validate-atlas, include = FALSE}
stopifnot(
  inherits(atlas, "atlas"),
  nrow(meta) == length(atlas$ids),
  identical(meta$id, atlas$ids),
  all(c("left", "right") %in% meta$hemi),
  length(atlas$ids) == 17L
)
```

Use `list_atlases()` to discover the other registered families without loading
them:

```{r discover-atlases}
list_atlases()[c("id", "label", "representation", "default_space")]
```

## How do you select regions?

Use `filter_atlas()` when the selection is naturally expressed through
metadata. Multiple expressions are intersected.

```{r filter-regions}
left_atlas <- filter_atlas(atlas, hemi == "left")

roi_metadata(left_atlas)[c("id", "label", "hemi")]
```

The result is still an atlas, now containing only the seven selected regions.
Excluded voxels are unlabelled; region IDs are preserved.

Atlas-level subsetting currently applies to volume atlases. For a `surfatlas`,
both `filter_atlas()` and `sub_atlas()` fail with a typed unsupported-operation
error. Use `get_roi()` to extract labelled surface ROIs instead.

```{r validate-filter, include = FALSE}
left_meta <- roi_metadata(left_atlas)
stopifnot(
  inherits(left_atlas, "atlas"),
  length(left_atlas$ids) == 7L,
  all(left_meta$hemi == "left")
)
```

`get_roi()` answers a different question: it extracts the voxel set belonging
to a named region. Labels repeat across hemispheres, so specify `hemi` when you
need one side.

```{r extract-roi}
hippocampus <- get_roi(atlas, label = "Hippocampus", hemi = "left")
hippocampus
```

The return value is a named list because one query can legitimately return
several ROIs. Here it contains one `neuroim2::ROIVol`.

```{r validate-roi, include = FALSE}
stopifnot(
  identical(names(hippocampus), "Hippocampus"),
  methods::is(hippocampus[[1]], "ROIVol"),
  length(hippocampus[[1]]) > 0L
)
```

## How do you summarise an image by parcel?

The atlas and image must describe the same voxel grid. To make that contract
visible, this example constructs a deterministic left-to-right gradient in the
atlas space. Replace `image` with a `NeuroVol` read from your own analysis once
you have checked its space and dimensions.

```{r make-image}
atlas_space <- neuroim2::space(atlas$atlas)
atlas_dim <- dim(atlas$atlas)

x_gradient <- array(
  rep(seq_len(atlas_dim[1]), atlas_dim[2] * atlas_dim[3]),
  dim = atlas_dim
)
x_gradient <- as.numeric(scale(x_gradient))
x_gradient <- array(x_gradient, dim = atlas_dim)

image <- neuroim2::NeuroVol(x_gradient, atlas_space)
```

`reduce_atlas()` applies a statistic within every labelled parcel. A 3D image
returns a long tibble by default.

```{r reduce-parcels}
parcel_summary <- reduce_atlas(atlas, image, mean)
parcel_summary
```

The `region` column uses the atlas's full region labels. The values are finite
and vary across parcels because the input image varies in space.

```{r validate-reduction, include = FALSE}
stopifnot(
  identical(names(parcel_summary), c("region", "value")),
  nrow(parcel_summary) == length(atlas$ids),
  all(is.finite(parcel_summary$value)),
  stats::sd(parcel_summary$value) > 0.1
)
```

## What does `map_atlas()` return?

`map_atlas()` pairs one value per region with the atlas's labels and
hemispheres. It returns a tibble; it does **not** mutate the atlas or create a
new surface object.

```{r map-values}
mapped_values <- map_atlas(atlas, parcel_summary$value)
mapped_values
```

This table is useful for reporting or aggregation because it keeps each value
beside its label and hemisphere. Here we make the midline group explicit and
then compute a hemisphere-level summary:

```{r summarize-hemispheres}
mapped_values$hemi[is.na(mapped_values$hemi)] <- "midline"
hemisphere_summary <- stats::aggregate(
  statistic ~ hemi,
  data = mapped_values,
  FUN = mean
)
hemisphere_summary
```

The left/right sign difference is an executable check on the orientation of
our synthetic gradient, not a claim about brain organisation.

```{r validate-mapping, include = FALSE}
stopifnot(
  identical(mapped_values$label, meta$label_full),
  all(is.finite(mapped_values$statistic)),
  hemisphere_summary$statistic[hemisphere_summary$hemi == "left"] < 0,
  hemisphere_summary$statistic[hemisphere_summary$hemi == "right"] > 0
)
```

## How do you inspect the atlas spatially?

`plot()` renders a volume atlas with one discrete colour per region. Use a
montage to see coverage across slices or `view = "ortho"` for three intersecting
planes.

```{r plot-atlas, fig.alt = "Axial montage of the bundled FreeSurfer ASEG atlas, showing discrete subcortical and midline regions in distinct colours across six slices."}
plot(atlas, nslices = 6)
```

For a cortical `surfatlas`, pass one value per region directly to
`plot_brain()`. The surface articles develop that separate object flow. Verify
region IDs rather than assuming unrelated volume and surface products align by
shape or region count.

## When is resampling safe?

Resampling changes a voxel grid. It does not, by itself, establish anatomical
correspondence between different templates.

- Within one template space, nearest-neighbour resampling is appropriate for
  discrete atlas labels when you need a different resolution or grid.
- Between template spaces, use a validated spatial transform. An affine grid
  reslice is not a substitute for a nonlinear template-to-template warp.

The transform planner makes the distinction explicit:

```{r transform-plan}
atlas_transform_plan(
  "MNI152NLin6Asym",
  "MNI152NLin2009cAsym",
  data_type = "voxel"
)
```

At the time this article was built, that route is recorded as `planned`, not
available. Do not interpret `outspace = "MNI152NLin2009cAsym"` as a certified
nonlinear transformation until the plan reports an available backend.

```{r validate-transform-status, include = FALSE}
cross_template_plan <- atlas_transform_plan(
  "MNI152NLin6Asym",
  "MNI152NLin2009cAsym",
  data_type = "voxel"
)
stopifnot(cross_template_plan$status %in% c("available", "planned"))
```

## Where should you go next?

You now have the package's central path:

`atlas` + compatible `NeuroVol` -> regional tibble -> analysis or volume plot.

Continue with:

1. `vignette("atlas-visualization")` for volume figures and ROI colours.
2. `vignette("surface-parcellations")` for the surface object model and
   parcel-level values.
3. `vignette("surface-panels")` for publication-style comparison figures.
4. `vignette("working-with-templateflow")` when you need external template
   assets and understand their space contract.

The dilation and Wang visual-cortex articles are specialist workflows. They are
not prerequisites for ordinary ROI analysis.
