---
title: "Identify, cite, and trace an atlas"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: |
  %\VignetteIndexEntry{Identify, cite, and trace an atlas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
library(neuroatlas)
```

An atlas object should tell you which resource you loaded, where its data came
from, which coordinates it uses, and what to cite. That information stays with
the object when you save it. Inspecting it does not contact a server.

## Inspect a loaded atlas

The bundled ASEG atlas makes a small, offline example:

```{r inspect}
a <- get_aseg_atlas()
atlas_metadata(a)
```

The summary separates the atlas's identity from its current voxel geometry.
`MNI152_unspecified` means that the precise anatomical template has not been
verified. The voxel size and affine can still be measured from the data.
Matching a familiar image's dimensions is insufficient evidence to name its
registration target.

The complete record has a stable schema:

| Component | Contents |
|:--|:--|
| `identity` | Identifier, name, description, published version, species, coverage |
| `content` | Volume or surface, labels or probabilities, variant parameters, region count |
| `spatial` | Anatomical template, evidence basis, actual grid or mesh geometry |
| `citations` | Bibliographic records with atlas, distribution, template, or software roles |
| `artifacts` | Source files, release identifiers, license information, checksums |
| `history` | Operations, structured parameters, and neuroatlas version |
| `parents` | Metadata snapshots for composite and derived resources |

```{r geometry}
m <- atlas_metadata(a)
m$spatial$voxel_size
m$spatial$affine
```

Missing information is stored as `NA` and shown as "not recorded". A paper's
publication year is not a data release version. Likewise, a source URL is not
proof that downloaded bytes match a particular release.

## Get the right references

```{r citations}
atlas_citations(a)
utils::toBibtex(atlas_citations(a))
```

These are ordinary R `bibentry` objects. Use `role = "atlas"` to select the
atlas publications, or inspect the citation table to see each reference's role:

```{r citation-roles}
m$citations[, c("role", "title", "doi")]
```

A projected atlas may require the original atlas publication, the distributed
projection, and the reference template. For example, Glasser surface metadata
include the original Glasser publication and the Mills Figshare projection,
alongside references captured for the loaded TemplateFlow geometry.

The catalog uses the distributor's recommended references where available,
including [FSL's atlas references](https://fsl.fmrib.ox.ac.uk/fsl/docs/other/datasets.html).
Olsen's original publication remains explicitly unverified. References without
complete bibliographic fields remain title/URL entries; the accessor does not
guess the missing fields or resolve them online.

## Trace changes to the object

Filtering keeps the original source records and adds a history step with the
retained region identifiers:

```{r subset}
left <- filter_atlas(a, hemi == "left")
atlas_metadata(left)$content$regions
atlas_history(left)[, c("step", "action", "details")]
tail(atlas_history(left)$parameters, 1)
```

Resampling changes the current grid while preserving the original artifact's
resolution. Here the atlas is sampled at 2 mm:

```{r resample}
target <- neuroim2::NeuroSpace(
  dim = as.integer(ceiling(dim(a$atlas) / 2)),
  spacing = c(2, 2, 2),
  origin = neuroim2::origin(neuroim2::space(a$atlas))
)
coarse <- get_aseg_atlas(outspace = target)
atlas_ref(coarse)$resolution
atlas_artifacts(coarse)$resolution
tail(atlas_history(coarse)$parameters, 1)[[1]]$interpolation
```

Grid resampling does not perform anatomical registration. The native template
identity is retained; a requested target grid is recorded separately as
`spatial$sampling_reference` and in the resampling parameters. Supplying a
different TemplateFlow name as `outspace` does not prove that a nonlinear warp
to that template was applied.

`dilate_atlas()` records its radius and neighbour limit. `merge_atlases()` keeps
both parent metadata records and their references, records the identifier map
and overlap rule, and rejects conflicting grids or known template identities.
Surface subsetting remains unsupported; its existing diagnostic is unchanged.

## Templates and probability maps

Template volumes and surfaces use the same schema, attached as an attribute so
their `NeuroVol` and `SurfaceGeometry` classes remain intact:

```{r templates, eval=FALSE}
t <- get_template("MNI152NLin2009cAsym", resolution = "01")
template_metadata(t)
template_citations(t)

s <- load_surface_template("fsaverage", "pial", hemi = "both", density = "164k")
template_metadata(s)$spatial$vertex_count
template_metadata(s$L)$artifacts
```

These examples require TemplateFlow data and are not run when building this
article. The loader captures the resolved filename and query, template
description, available license and references, and actual geometry. A query's
resolution key, such as `"01"`, is retained as a parameter; voxel sizes come from
the image rather than interpreting that key as millimetres.
`TemplateFlowVersion` is retained separately in provenance; it is not treated
as the original template's published version.

`get_wang_prob_atlas(path_only = FALSE)` also provides `atlas_metadata()` and
`atlas_citations()`. Its record distinguishes probability maps from
maximum-probability labels, and retains the individual volume records in
`parents`. Path-only requests retain their existing path/manifest return types.

## Save the evidence with your analysis

```{r save, eval=FALSE}
saveRDS(left, "left-aseg.rds")
restored <- readRDS("left-aseg.rds")
atlas_metadata(restored)
```

Artifact checksums describe source bytes, not the current in-memory atlas after
modifications. R versions with `tools::sha256sum()` record SHA-256; older R
versions record a labelled MD5 checksum. For installed bundled datasets,
`checksum_basis = "source_manifest"` identifies the checksum of the source
`.rda` file, which R may have converted to its lazy-load database. `read_file`
identifies a file hashed at load time. A recorded checksum is a content identity,
not an assertion that an upstream publisher supplied a matching checksum.

The metadata record is authoritative. Existing `atlas_ref()`,
`atlas_artifacts()`, and `atlas_history()` calls continue to work for atlas
objects. Direct edits to legacy compatibility fields do not update the record.
Old saved objects receive conservative metadata views with unknown provenance.
Preservation is supported for neuroatlas's loaders and atlas operations;
arbitrary data replacement or transformations in other packages may require
their own provenance handling.
