---
title: Composing Surface Panel Figures
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Composing Surface Panel Figures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>
---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(albersdown::theme_albers(params$family))
}

can_plot <- requireNamespace("neuroatlas", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE) &&
  requireNamespace("neurosurf", quietly = TRUE)

can_layout <- can_plot &&
  requireNamespace("patchwork", quietly = TRUE) &&
  requireNamespace("scico", quietly = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = can_layout
)
```

```{r}
library(neuroatlas)
```

## What problem are we solving?

You often have one or more parcel-level maps and need a figure that is
already arranged, labeled, and publication-ready. The awkward part is
usually not the parcel values themselves; it is the repeated patchwork
work afterward: moving the legend, shortening facet labels, keeping the
same scale across panels, and making sure left and right hemispheres are
labeled anatomically rather than by screen position.

`plot_brain()` and `plot_brain_grid()` are the high-level entry points
for that job. They let you compose static surface figures directly from
parcel values instead of building the figure by hand after plotting.

## What do you need before you start?

For panel figures, you need a surface atlas and one numeric value per
parcel.

```{r atlas-and-values}
atl <- schaefer_surf(
  parcels = 200,
  networks = 7,
  space = "fsaverage6",
  surf = "inflated"
)

parcel_vals <- seq(-2, 2, length.out = length(atl$ids))
```

For comparison figures, `plot_brain_grid()` takes a named list of those
numeric vectors.

```{r multiple-maps}
vals_list <- list(
  Baseline = sin(seq_along(atl$ids) / 15),
  Follow_up = cos(seq_along(atl$ids) / 18),
  Difference = sin(seq_along(atl$ids) / 15) -
    cos(seq_along(atl$ids) / 18)
)
```

## How do you build one polished panel figure?

Use `plot_brain()` when you want one figure with a few surface panels
and a single legend.

```{r single-figure, fig.width=8, fig.height=4.5, fig.alt="Surface panel figure with left and right hemisphere lateral and medial views plus a bottom colorbar"}
plot_brain(
  atl,
  vals = parcel_vals,
  views = c("lateral", "medial"),
  interactive = FALSE,
  style = "ggseg_like",
  colorbar = "bottom",
  colorbar_title = "Standardized effect",
  title = "Parcel-level summary on fsaverage6",
  subtitle = "Bottom colorbar plus concise panel labels",
  panel_labels = c(
    "Left Lateral" = "LH lateral",
    "Right Lateral" = "RH lateral",
    "Left Medial" = "LH medial",
    "Right Medial" = "RH medial"
  )
)
```

This is the default static workflow when you want:

- view-aware facet labels without post-hoc editing
- one explicit legend instead of an implicit colour mapping
- a single annotated figure object you can print or save directly

## How do you compare several maps with one legend?

Use `plot_brain_grid()` when each panel is a different map but the
colour scale should mean the same thing everywhere.

```{r grid-figure, fig.width=10, fig.height=7, fig.alt="Four-panel surface comparison figure with shared right-side colorbar and repeated hemisphere view labels"}
plot_brain_grid(
  atl,
  vals_list,
  views = c("lateral", "medial"),
  titles = c("Baseline", "Follow-up", "Difference"),
  shared_scale = TRUE,
  colorbar = "right",
  colorbar_title = "z-score",
  title = "Comparing parcel maps with a shared legend",
  subtitle = "Every panel uses the same limits and the same view labels",
  panel_labels = c(
    "Left Lateral" = "LH lateral",
    "Right Lateral" = "RH lateral",
    "Left Medial" = "LH medial",
    "Right Medial" = "RH medial"
  ),
  style = "ggseg_like",
  ncol = 2
)
```

The arguments that matter most here are:

- `titles`: names each map
- `shared_scale = TRUE`: enforces one global value range
- `colorbar`: places a shared legend on the right or bottom
- `panel_labels`: simplifies the repeated surface-view labels

If you want each panel to optimise its own within-panel contrast, use
`shared_scale = FALSE`. That can make weak patterns easier to see, but
you lose strict colour comparability across panels.

## What do the default panel labels mean?

The default panel labels are anatomical hemisphere plus view:

- `Left Lateral`
- `Right Lateral`
- `Left Medial`
- `Right Medial`

Those names describe the surface itself, not where you happen to place
the panel in a larger composition. A right-hemisphere panel is still a
right hemisphere if you later move it to the left side of a page.

`plot_brain()` now uses a fixed anatomical convention for lateral and
medial views:

- left hemispheres are drawn with anterior on the left
- right hemispheres are drawn with anterior on the right

That means `panel_labels = c("Right Lateral" = "Right hemisphere")` is a
safe simplification when you are showing one view per hemisphere. If you
mix lateral and medial views, keep the view in the label unless you have
a strong reason to collapse it.

## Which layout controls are worth remembering?

These are the arguments you will use repeatedly:

- `style = "ggseg_like"` gives a presentation-oriented default layout
- `panel_layout = "presentation"` is the explicit layout control behind that style
- `panel_labels` rewrites facet labels without changing the underlying hemisphere/view mapping
- `colorbar` and `colorbar_title` control the standalone legend
- `title`, `subtitle`, and `caption` annotate the composed figure directly
- `ncol` controls the panel grid for both `plot_brain()` and `plot_brain_grid()`

If you need the raw projected coordinates for a custom workflow, use
`panel_layout = "native"`. If you want the cleaner, more balanced
publication layout, use `panel_layout = "presentation"` or
`style = "ggseg_like"`.

## Where should you go next?

Use this vignette as the reference for figure composition, then branch
out depending on what you need next:

- `vignette("surface-parcellations", package = "neuroatlas")` for loading surface atlases and overlays
- `vignette("atlas-visualization", package = "neuroatlas")` for volumetric plotting and palette workflows
- `?plot_brain` for the full single-figure API
- `?plot_brain_grid` for the shared-scale multi-panel API
