---
title: Composing Surface Panel Figures
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Composing Surface Panel Figures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
}

can_plot <- requireNamespace("neuroatlas", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE) &&
  requireNamespace("neurosurf", quietly = TRUE)

can_layout <- can_plot &&
  requireNamespace("patchwork", quietly = TRUE) &&
  requireNamespace("scico", quietly = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = can_layout
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```


```{r}
library(neuroatlas)
```

## What problem are we solving?

You often have either parcel-level estimates or a continuous statistic on the
cortical surface and need a figure that is already arranged, labelled, and
publication-ready. The awkward part is not only panel composition. The figure
must also preserve what the values mean: parcel maps may show atlas boundaries,
whereas a continuous field should appear on a quiet anatomical substrate with
no parcel or occlusion lines drawn over it.

`plot_brain()` and `plot_brain_grid()` are the high-level entry points
for that job. They let you compose static surface figures directly from
parcel values instead of building the figure by hand after plotting.

## What do you need before you start?

For a parcel figure, you need a surface atlas and one numeric value per parcel.
For a continuous figure, you use the same atlas to supply geometry, cortex
mask, and anatomical provenance, then provide one value per surface vertex (or
a `NeuroVol` that can be projected to the surface).

```{r atlas-and-values}
atl <- schaefer_surf(
  parcels = 200,
  networks = 7,
  space = "fsaverage6",
  surf = "inflated"
)

parcel_vals <- seq(-2, 2, length.out = length(atl$ids))
```

For comparison figures, `plot_brain_grid()` takes a named list of those
numeric vectors.

```{r multiple-maps}
vals_list <- list(
  Baseline = sin(seq_along(atl$ids) / 15),
  Follow_up = cos(seq_along(atl$ids) / 18),
  Difference = sin(seq_along(atl$ids) / 15) -
    cos(seq_along(atl$ids) / 18)
)
```

## How do you build one polished panel figure?

Use `plot_brain()` when you want one figure with a few surface panels
and a single legend.

```{r single-figure, fig.width=8, fig.height=5, fig.alt="Surface panel figure with left and right hemisphere lateral and medial views plus a bottom colorbar"}
plot_brain(
  atl,
  vals = parcel_vals,
  views = c("lateral", "medial"),
  interactive = FALSE,
  style = "ggseg_like",
  colorbar = "bottom",
  colorbar_title = "Standardized effect",
  title = "Parcel-level summary on fsaverage6",
  subtitle = "Bottom colorbar plus concise panel labels",
  panel_labels = c(
    "Left Lateral" = "LH lateral",
    "Right Lateral" = "RH lateral",
    "Left Medial" = "LH medial",
    "Right Medial" = "RH medial"
  )
)
```

This is the default static workflow when you want:

- view-aware facet labels without post-hoc editing
- one explicit legend instead of an implicit colour mapping
- a single annotated figure object you can render or save directly

## How do you show a continuous statistic instead?

Use `style = "stat_publication"` when the colour represents a continuous
vertex-wise field rather than parcels. The preset selects a neutral cortex,
removes parcel and culling-derived silhouette lines, keeps anatomy below the
statistic, and obtains the legend from the overlay rather than from a dummy
base map.

This example derives a smooth synthetic field from the inflated coordinates.
Replace `stat_overlay` with your own `list(lh = ..., rh = ...)`, or pass a
`NeuroVol` as `overlay` and let `plot_brain()` project it.

```{r continuous-overlay-data, include=FALSE}
surface_statistic <- function(geometry) {
  xyz <- neurosurf::vertices(geometry)
  value <- sin(xyz[, 2] / 20) + cos(xyz[, 3] / 16)
  as.numeric(scale(value))
}

stat_overlay <- list(
  lh = surface_statistic(atl$lh_atlas@geometry),
  rh = surface_statistic(atl$rh_atlas@geometry)
)

stopifnot(
  length(stat_overlay$lh) == nrow(neurosurf::vertices(atl$lh_atlas@geometry)),
  length(stat_overlay$rh) == nrow(neurosurf::vertices(atl$rh_atlas@geometry)),
  all(is.finite(unlist(stat_overlay))),
  stats::sd(stat_overlay$lh) > 0,
  stats::sd(stat_overlay$rh) > 0
)
```

```{r continuous-figure, fig.width=10, fig.height=6.25, fig.alt="Four-view cortical surface figure showing a continuous signed statistic over subtle anatomy, with a neutral medial wall, orientation marks, and an overlay-derived bottom colorbar"}
stat_figure <- plot_brain(
  atl,
  overlay = stat_overlay,
  overlay_threshold = 1,
  overlay_lim = c(-2.5, 2.5),
  overlay_title = "z-score",
  style = "stat_publication",
  static_backend = "cpu",
  interactive = FALSE,
  colorbar = "bottom",
  medial_wall = "shade",
  camera = "canonical",
  title = "Continuous statistic on fsaverage6"
)

stat_figure
```

```{r validate-continuous-figure, include=FALSE}
legend_meta <- attr(stat_figure, "plot_brain_colorbar")
anatomy_meta <- attr(stat_figure, "plot_brain_anatomy")
stopifnot(
  identical(attr(stat_figure, "plot_brain_backend"), "cpu_barycentric"),
  identical(legend_meta$source, "overlay"),
  identical(legend_meta$lim, c(-2.5, 2.5)),
  identical(legend_meta$breaks, c(-2.5, -1, 0, 1, 2.5)),
  anatomy_meta$anatomy$lh$topology_verified,
  anatomy_meta$anatomy$rh$topology_verified
)
```

Notice that the threshold appears as marks at `-1` and `1`, while the colour
limits remain exactly `[-2.5, 2.5]`. The cortex mask is independent of parcel
values, so the medial wall cannot acquire overlay colour merely because an
atlas uses label zero. Curvature is resolved from matched white geometry and
transferred only after vertex correspondence is checked.

When `overlay` is a volume, the publication preset uses trilinear sampling at
five fractions from 0.1 through 0.9 of the white-to-pial ribbon, averages those
depth samples, and applies no tangential smoothing unless you request it. The
effective projection, mask, anatomy, camera, and legend sources are retained as
attributes on the returned figure.

## How do you compare several maps with one legend?

Use `plot_brain_grid()` when each panel is a different map but the
colour scale should mean the same thing everywhere.

```{r grid-figure, fig.width=10, fig.height=7, fig.alt="Four-panel surface comparison figure with shared right-side colorbar and repeated hemisphere view labels"}
plot_brain_grid(
  atl,
  vals_list,
  views = c("lateral", "medial"),
  titles = c("Baseline", "Follow-up", "Difference"),
  shared_scale = TRUE,
  colorbar = "right",
  colorbar_title = "z-score",
  title = "Comparing parcel maps with a shared legend",
  subtitle = "Every panel uses the same limits and the same view labels",
  panel_labels = c(
    "Left Lateral" = "LH lateral",
    "Right Lateral" = "RH lateral",
    "Left Medial" = "LH medial",
    "Right Medial" = "RH medial"
  ),
  style = "ggseg_like",
  ncol = 2
)
```

The arguments that matter most here are:

- `titles`: names each map
- `shared_scale = TRUE`: enforces one global value range
- `colorbar`: places a shared legend on the right or bottom
- `panel_labels`: simplifies the repeated surface-view labels

If you want each panel to optimise its own within-panel contrast, use
`shared_scale = FALSE`. That can make weak patterns easier to see, but
you lose strict colour comparability across panels.

## How should you label the panels?

Panel labels name the surface being shown: hemisphere plus view, such as
`Left Lateral` or `Right Medial`. They do not describe where the panel
sits in the final grid.

In lateral and medial views, `plot_brain()` uses a fixed anatomical
orientation: left hemispheres have anterior on the left, and right
hemispheres have anterior on the right.

Use `panel_labels` when the default labels are too long. Keep both the
hemisphere and the view when a figure mixes lateral and medial views; use
shorter labels such as `LH` and `RH` only when each hemisphere appears in
one view.

## What should you remember?

Choose the style from the scientific object you are displaying:

- use `style = "ggseg_like"` for parcel identities or parcel-valued estimates;
- use `style = "stat_publication"` for a continuous vertex field or projected
  statistic volume;
- keep the CPU backend for deterministic static reports and use the existing
  interactive route for atlas exploration and hover behaviour.

Then adjust:

- `panel_labels` for shorter facet labels
- `colorbar` and `colorbar_title` for the shared legend
- `title`, `subtitle`, and `caption` for figure annotations
- `ncol` for the panel grid

For a continuous field, prefer `overlay_threshold`, `overlay_lim`,
`medial_wall`, and `camera` over low-level line-width controls. Parcel
boundaries and internal depth discontinuities are not activation contours.

Use `panel_layout = "native"` only when you need the raw projected
coordinates for a custom workflow.

## Where should you go next?

Use this vignette as the reference for figure composition, then branch
out depending on what you need next:

- `vignette("surface-parcellations", package = "neuroatlas")` for loading surface atlases and overlays
- `vignette("atlas-visualization", package = "neuroatlas")` for volumetric plotting and palette workflows
- `?plot_brain` for the full single-figure API
- `?plot_brain_grid` for the shared-scale multi-panel API
