---
title: Composing Surface Panel Figures
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{Composing Surface Panel Figures} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
}

can_plot <- requireNamespace("neuroatlas", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE) &&
  requireNamespace("neurosurf", quietly = TRUE)

can_layout <- can_plot &&
  requireNamespace("patchwork", quietly = TRUE) &&
  requireNamespace("scico", quietly = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = can_layout
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

```{r}
library(neuroatlas)
```

## What problem are we solving?

You often have one or more parcel-level maps and need a figure that is
already arranged, labeled, and publication-ready. The awkward part is
usually not the parcel values themselves; it is the repeated patchwork
work afterward: moving the legend, shortening facet labels, keeping the
same scale across panels, and making sure left and right hemispheres are
labeled anatomically rather than by screen position.

`plot_brain()` and `plot_brain_grid()` are the high-level entry points
for that job. They let you compose static surface figures directly from
parcel values instead of building the figure by hand after plotting.

## What do you need before you start?

For panel figures, you need a surface atlas and one numeric value per
parcel.

```{r atlas-and-values}
atl <- schaefer_surf(
  parcels = 200,
  networks = 7,
  space = "fsaverage6",
  surf = "inflated"
)

parcel_vals <- seq(-2, 2, length.out = length(atl$ids))
```

For comparison figures, `plot_brain_grid()` takes a named list of those
numeric vectors.

```{r multiple-maps}
vals_list <- list(
  Baseline = sin(seq_along(atl$ids) / 15),
  Follow_up = cos(seq_along(atl$ids) / 18),
  Difference = sin(seq_along(atl$ids) / 15) -
    cos(seq_along(atl$ids) / 18)
)
```

## How do you build one polished panel figure?

Use `plot_brain()` when you want one figure with a few surface panels
and a single legend.

```{r single-figure, fig.width=8, fig.height=5, fig.alt="Surface panel figure with left and right hemisphere lateral and medial views plus a bottom colorbar"}
plot_brain(
  atl,
  vals = parcel_vals,
  views = c("lateral", "medial"),
  interactive = FALSE,
  style = "ggseg_like",
  colorbar = "bottom",
  colorbar_title = "Standardized effect",
  title = "Parcel-level summary on fsaverage6",
  subtitle = "Bottom colorbar plus concise panel labels",
  panel_labels = c(
    "Left Lateral" = "LH lateral",
    "Right Lateral" = "RH lateral",
    "Left Medial" = "LH medial",
    "Right Medial" = "RH medial"
  )
)
```

This is the default static workflow when you want:

- view-aware facet labels without post-hoc editing
- one explicit legend instead of an implicit colour mapping
- a single annotated figure object you can render or save directly

## How do you compare several maps with one legend?

Use `plot_brain_grid()` when each panel is a different map but the
colour scale should mean the same thing everywhere.

```{r grid-figure, fig.width=10, fig.height=7, fig.alt="Four-panel surface comparison figure with shared right-side colorbar and repeated hemisphere view labels"}
plot_brain_grid(
  atl,
  vals_list,
  views = c("lateral", "medial"),
  titles = c("Baseline", "Follow-up", "Difference"),
  shared_scale = TRUE,
  colorbar = "right",
  colorbar_title = "z-score",
  title = "Comparing parcel maps with a shared legend",
  subtitle = "Every panel uses the same limits and the same view labels",
  panel_labels = c(
    "Left Lateral" = "LH lateral",
    "Right Lateral" = "RH lateral",
    "Left Medial" = "LH medial",
    "Right Medial" = "RH medial"
  ),
  style = "ggseg_like",
  ncol = 2
)
```

The arguments that matter most here are:

- `titles`: names each map
- `shared_scale = TRUE`: enforces one global value range
- `colorbar`: places a shared legend on the right or bottom
- `panel_labels`: simplifies the repeated surface-view labels

If you want each panel to optimise its own within-panel contrast, use
`shared_scale = FALSE`. That can make weak patterns easier to see, but
you lose strict colour comparability across panels.

## How should you label the panels?

Panel labels name the surface being shown: hemisphere plus view, such as
`Left Lateral` or `Right Medial`. They do not describe where the panel
sits in the final grid.

In lateral and medial views, `plot_brain()` uses a fixed anatomical
orientation: left hemispheres have anterior on the left, and right
hemispheres have anterior on the right.

Use `panel_labels` when the default labels are too long. Keep both the
hemisphere and the view when a figure mixes lateral and medial views; use
shorter labels such as `LH` and `RH` only when each hemisphere appears in
one view.

## What should you remember?

For most figures, start with `style = "ggseg_like"`. Then adjust:

- `panel_labels` for shorter facet labels
- `colorbar` and `colorbar_title` for the shared legend
- `title`, `subtitle`, and `caption` for figure annotations
- `ncol` for the panel grid

Use `panel_layout = "native"` only when you need the raw projected
coordinates for a custom workflow.

## Where should you go next?

Use this vignette as the reference for figure composition, then branch
out depending on what you need next:

- `vignette("surface-parcellations", package = "neuroatlas")` for loading surface atlases and overlays
- `vignette("atlas-visualization", package = "neuroatlas")` for volumetric plotting and palette workflows
- `?plot_brain` for the full single-figure API
- `?plot_brain_grid` for the shared-scale multi-panel API
