---
title: Verified Template Transforms
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2
    css: albers.css
    includes:
      in_header: albers-header.html
resource_files:
- albers.css
- albers-header.html
- fonts
vignette: |
  %\VignetteIndexEntry{Verified Template Transforms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE)
```

A template transform moves samples between explicitly named image grids. It
does not estimate a registration, infer an atlas correspondence, or change a
parcellation's region identities. `neuroatlas` resolves an available route and
samples the source once on the target grid.

The application API requires the optional `neurotransform` and `hdf5r` packages.
Install the engine revision pinned in neuroatlas's `Remotes` field; the runtime
checks its H5 conventions against independent fixtures.

## Transform a volume on an explicit target grid

Use `get_template_transform()` followed by `apply_template_transform()`.
The built-in MNI305-to-MNI152 affine route is available without a download.
This offline example supplies continuous semantics and both image grids.

```{r affine-workflow}
library(neuroatlas)
if (!requireNamespace("neurotransform", quietly = TRUE)) {
  stop("This vignette requires the optional neurotransform package.")
}

source_grid <- neuroim2::NeuroSpace(
  c(5L, 5L, 5L), spacing = c(2, 2, 2), origin = c(-4, -4, -4)
)
source <- neuroim2::NeuroVol(array(seq_len(125), c(5, 5, 5)), source_grid)
target_grid <- neuroim2::NeuroSpace(
  c(5L, 5L, 5L), spacing = c(2, 2, 2), origin = c(-4, -4, -4)
)

transform <- get_template_transform(
  "MNI305", "MNI152", provider = "neuroatlas", offline = TRUE
)
transformed <- apply_template_transform(
  source, transform, target_grid, data_type = "continuous"
)

c(
  source_range = range(as.vector(source)),
  transformed_range = range(as.vector(transformed)),
  output_dimensions = paste(dim(transformed), collapse = " x ")
)
```

The target grid is part of the analysis definition: dimensions, affine,
spacing, and origin determine where values are evaluated. A bare grid asserts
that it belongs to the transform's target template. Attached source or target
metadata must agree with the route.

```{r receipt}
receipt <- attr(transformed, "neuroatlas_transform")
receipt[c("from_space", "to_space", "data_type", "interpolation", "renormalized")]
```

## Select sampling semantics

`data_type` is a sampling contract. Labels and masks use nearest-neighbour
sampling, so output values remain source labels or zero outside the source
field. Continuous values and probability channels use linear interpolation.
Probability values are not renormalized after resampling.

For an atlas, label semantics are automatic. For a bare `NeuroVol` or
`NeuroVec`, supply `data_type` unless metadata already declares its contents.
The package rejects a conflict between metadata and an explicit type.

When a target clips a labelled atlas, the atlas retains all semantic `ids`,
labels, and provenance. The receipt records only IDs absent from the output.

```{r lost-labels, eval=FALSE}
aligned_atlas <- transform_atlas(
  atlas, to_space = "MNI152", target = target_grid, provider = "neuroatlas"
)
attr(aligned_atlas, "neuroatlas_transform")$lost_label_ids
```

`transform_atlas()` changes image space; it does not make labels from different
parcellations equivalent. After alignment, use `atlas_overlap()` to inspect
spatial relationships. `map_atlas()` retains its original role of mapping
supplied regional values onto an atlas.

## Route availability and nonlinear artifacts

The manifest distinguishes available and planned routes. Only available routes
are resolved by `get_template_transform()`.

```{r manifest}
space_transform_manifest()[, c("from_space", "to_space", "backend", "status")]
```

The MNI152NLin6Asym-to-MNI152NLin2009cAsym nonlinear route is currently
planned. The neuroatlas-built ANTs H5 pair requires a published immutable URL,
checksum, and passing qualification record. This is the intended
call after that release becomes available, so it is deliberately not evaluated.

```{r nonlinear-release, eval=FALSE}
nonlinear <- get_template_transform(
  "MNI152NLin6Asym", "MNI152NLin2009cAsym", provider = "neuroatlas",
  cache_dir = transform_cache_path(), download = TRUE, offline = FALSE
)
```

## Work with the verified cache

Artifacts live in a dedicated cache, separate from TemplateFlow's cache. A
download is published atomically only after its byte count and SHA-256 digest
match its immutable release record. Verification cannot be disabled; a corrupt
cache entry fails rather than being used silently.

Use `offline = TRUE` when an analysis may use only already verified artifacts:

```{r offline-cache, eval=FALSE}
get_template_transform(
  "MNI152NLin6Asym", "MNI152NLin2009cAsym", provider = "neuroatlas",
  cache_dir = transform_cache_path(), offline = TRUE
)
```

`clear_transform_cache()` removes only entries neuroatlas recognizes as its own,
refuses active locks, and does not clear a TemplateFlow cache.

```{r clear-cache, eval=FALSE}
clear_transform_cache(artifact_version = "released-artifact-version")
```

Keep the `neuroatlas_transform` receipt with analysis output. It records the
route, artifact checksums, grids, interpolation, and lost-label information.
