---
title: Working with TemplateFlow in neuroatlas
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Working with TemplateFlow in neuroatlas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "ragg_png")
}
if (requireNamespace("systemfonts", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  albersdown::albers_register_fonts()
}
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(
    albersdown::theme_albers(
      family = params$family,
      preset = params$preset
    )
  )
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```


[TemplateFlow](https://www.templateflow.org/) gives neuroimaging workflows a
shared, queryable archive of templates and related assets. The question is
usually specific: *which file describes this tissue, surface, or atlas in this
template space and resolution?* `neuroatlas` translates that query into either
a file path or a `neuroim2::NeuroVol`.

The current backend is the pure-R `templateflow` package. Python, reticulate,
and a connection object are not part of the supported setup.

## What do you need before you start?

Install `neuroatlas` with its `templateflow` import. You can then check the
backend and inspect its disk-cache location without downloading an image:

```{r load}
library(neuroatlas)

requireNamespace("templateflow", quietly = TRUE)
show_templateflow_cache_path()
```

`create_templateflow()` and `install_templateflow()` are deprecated migration
stubs. New code should call `get_template()`, `tflow_spaces()`, and
`tflow_files()` directly.

## How do you fetch one template image?

This is the ordinary path:

```{r fetch-brain, eval = FALSE}
brain <- get_template(
  space = "MNI152NLin2009cAsym",
  variant = "brain",
  modality = "T1w",
  resolution = 2
)
brain
```

The result is a `NeuroVol`: voxel values plus dimensions, spacing, origin, and
affine geometry. If another tool should read the file, request the path instead:

```{r fetch-path, eval = FALSE}
brain_path <- get_template(
  space = "MNI152NLin2009cAsym",
  variant = "brain",
  modality = "T1w",
  resolution = 2,
  path_only = TRUE
)
brain_path
```

These chunks are not run during package builds because the first call may use
the network and write to the TemplateFlow cache. They are intentionally shown
without fabricated output.

## How do the query fields fit together?

TemplateFlow filenames use BIDS-like entities. `get_template()` exposes two
levels of control:

| Argument | Meaning | Example |
|---|---|---|
| `space` | template coordinate system | `"MNI152NLin2009cAsym"` |
| `resolution` | volumetric resolution in millimetres | `2` |
| `variant` | convenient high-level asset type | `"brain"`, `"mask"`, `"probseg"` |
| `modality` | image suffix when applicable | `"T1w"`, `"T2w"` |
| `label` | tissue or structure label | `"GM"`, `"WM"`, `"CSF"` |
| `atlas` | atlas entity | `"Schaefer2018"` |
| `desc`, `suffix` | explicit low-level entities | `"brain"`, `"dseg"` |

Use the high-level `variant` interface for common images:

```{r common-assets, eval = FALSE}
brain_mask <- get_template(
  "MNI152NLin2009cAsym",
  variant = "mask",
  resolution = 2
)

gm_probability <- get_template(
  "MNI152NLin2009cAsym",
  variant = "probseg",
  label = "GM",
  resolution = 2
)
```

Use explicit entities when the archive contains a specialised asset:

```{r atlas-asset, eval = FALSE}
schaefer_path <- get_template(
  space = "MNI152NLin2009cAsym",
  atlas = "Schaefer2018",
  desc = "100Parcels7Networks",
  suffix = "dseg",
  resolution = 2,
  path_only = TRUE
)
```

If one argument has several values, `get_template()` returns a named list:

```{r vectorized, eval = FALSE}
templates <- get_template(
  space = "MNI152NLin2009cAsym",
  variant = "brain",
  resolution = c(1, 2),
  path_only = TRUE
)
names(templates)
```

Only one parameter may be vectorised in a call. For a grid of queries, iterate
explicitly so every requested combination is visible in your code.

## How do you discover what exists?

`tflow_spaces()` lists registered template IDs; its optional regular expression
filters the result:

```{r list-spaces, eval = FALSE}
tflow_spaces(pattern = "^MNI")
```

`tflow_files()` is the lower-level inventory query. It returns paths matching
the supplied entities and is useful before you choose one exact file:

```{r list-files, eval = FALSE}
tflow_files(
  "MNI152NLin2009cAsym",
  query_args = list(
    suffix = "T1w",
    resolution = 2
  )
)
```

Archive contents evolve independently of this package. Discover fields from
the live inventory rather than treating a vignette's example roster as a
permanent catalogue.

## How do you fetch surface geometry?

Surface queries use hemisphere, density or resolution, and a surface suffix.
`get_surface_template()` returns a file path by default:

```{r surface-path, eval = FALSE}
left_midthickness <- get_surface_template(
  template_id = "fsLR",
  surface_type = "midthickness",
  hemi = "L",
  density = "32k"
)
left_midthickness
```

`load_surface_template()` reads the geometry for use with `neurosurf`:

```{r surface-object, eval = FALSE}
geometry <- load_surface_template(
  template_id = "fsLR",
  surface_type = "midthickness",
  hemi = "L",
  density = "32k"
)
geometry
```

Use `vignette("surface-templates")` for the distinction between geometry,
per-vertex data, and labelled surface atlases.

## Does a matching grid mean a matching space?

No. A TemplateFlow identifier describes an anatomical template, not merely an
array shape. Two images can have equal dimensions and still require a spatial
transform; two grids can differ while remaining in the same template space.

Before combining an atlas and another image, check both:

1. Do their declared template spaces match?
2. Do their voxel grids match?

If the template spaces differ, consult the transform plan before resampling:

```{r transform-contract}
atlas_transform_plan(
  "MNI152NLin6Asym",
  "MNI152NLin2009cAsym",
  data_type = "voxel"
)
```

```{r validate-transform-contract, include = FALSE}
plan <- atlas_transform_plan(
  "MNI152NLin6Asym",
  "MNI152NLin2009cAsym",
  data_type = "voxel"
)
stopifnot(plan$status %in% c("available", "planned"))
```

A plan marked `planned` is not an executable transform. Do not use ordinary
grid resampling to imply nonlinear registration between templates.

## What is cached?

The pure-R `templateflow` package manages downloaded files on disk. Within an R
session, `neuroatlas` also memoises loading a resolved NIfTI path into a
`NeuroVol`.

```{r cache}
show_templateflow_cache_path()
```

`clear_templateflow_cache()` removes cached assets and in-memory loading state;
because that is destructive and may trigger large downloads later, call it
only when you deliberately want to rebuild the cache.

## Where should you go next?

- Return to `vignette("neuroatlas-overview")` for the atlas-to-summary workflow.
- Read `vignette("surface-templates")` for surface geometry and data.
- Use `space_transform_manifest()` and `atlas_transform_plan()` before mixing
  images from different template spaces.

For archive naming and governance, consult the
[TemplateFlow documentation](https://www.templateflow.org/).
