---
title: "Working with TemplateFlow in neuroatlas"
author: "neuroatlas Dev Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with TemplateFlow in neuroatlas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  eval = FALSE # Default to FALSE for most chunks to avoid downloads during vignette build
               # We can selectively set eval = TRUE for specific chunks
)
library(neuroatlas)
```

## Introduction

[TemplateFlow](https://www.templateflow.org/) is a repository of curated neuroimaging templates and atlases, accessible via a Python client. The `neuroatlas` package provides an R interface to TemplateFlow, allowing you to easily discover, download, and use these resources within your R workflows.

This vignette provides a guide to using the TemplateFlow functionalities in `neuroatlas`.

## Getting Started: The `templateflow` Object

The main entry point for interacting with TemplateFlow is by creating a `templateflow` interface object. This object manages the connection to the Python API and caching configurations.

```{r create_tf_object, eval = FALSE}
# Create a templateflow interface object
tf <- create_templateflow()
print(tf)
```

If Python or the templateflow module are not available, you'll need to install them first:

```{r install_tf, eval = FALSE}
# Install the Python templateflow package
install_templateflow()
```

Running `print(tf)` (if `tf` is successfully created) will show you:
- The local cache path where TemplateFlow files will be stored. `neuroatlas` sets this up in a user-specific directory (e.g., `~/.cache/neuroatlas/templateflow` on Linux).
- The status of the connection to the Python API.
- A few examples of available templates and their total count.

Output might look like:
```
#> <neuroatlas TemplateFlow Interface>
#>   Cache Path:  /Users/youruser/Library/Caches/org.R-project.R/R/neuroatlas/templateflow 
#>   API Status:  Connected (Python API handle initialized) 
#>   Available Templates (Examples):  AMYGDALA, ARA, BRODMANN, CHARITE, Fischer344, ... (Total: 76)
```
(Note: The exact cache path and template list may vary.)

The underlying Python TemplateFlow library uses the `TEMPLATEFLOW_HOME` environment variable to determine its storage location. `neuroatlas` attempts to set this to its managed cache path to centralize downloads.

## Discovering Templates and Metadata

Before fetching files, you might want to see what's available.

### Listing Available Template Spaces

Use `tflow_spaces()` to get a list of all top-level template space identifiers:

```{r list_templates, eval = FALSE}
# List all available template spaces
all_templates <- tflow_spaces()
print(head(all_templates))
cat("Total templates available:", length(all_templates), "\n")
```
Example output:
```
#> [1] "AMYGDALA" "ARA"      "BRODMANN" "CHARITE"  "Fischer344" "GS"
#> Total templates available: 76
```

You can also filter this list with a pattern (regular expression):
```{r list_templates_pattern, eval = FALSE}
# Find all MNI templates
mni_templates <- tflow_spaces(pattern = "MNI")
print(mni_templates)
```
Example output:
```
#> [1] "MNI152Lin"           "MNI152NLin2009cAsym" "MNI152NLin6Asym"    
#> [4] "MNI152NLin6Sym"      "MNIInfant"           "MNIPediatricAsym"
```

### Listing Specific Files (Metadata Query)

To find specific files within a template space based on metadata (like suffix, resolution, description), use `tflow_files()`:

```{r list_metadata, eval = FALSE}
# Find all T1w images for the 'MNI152NLin2009cAsym' space
mni_t1w_files <- tflow_files(
  space = "MNI152NLin2009cAsym",
  query_args = list(suffix = "T1w")
)
print(head(mni_t1w_files))

# Find brain masks for 'MNI152NLin2009cAsym' at 1mm resolution
mni_brain_masks <- tflow_files(
  space = "MNI152NLin2009cAsym",
  query_args = list(suffix = "mask", desc = "brain", resolution = "1")
)
print(mni_brain_masks)
```
This function returns a character vector of full paths to the files that TemplateFlow would download/has cached.

## Fetching Templates with `get_template()`

The primary function for fetching template files is `get_template()`. It provides a flexible R interface to the TemplateFlow archive.

```{r get_template_setup, echo=FALSE, eval=TRUE}
# For vignette build, use a placeholder if we can't eval real calls
mock_neurovol <- function(name = "mock_vol") {
  vol <- list(name = name, data = array(0, dim=c(2,2,2)))
  class(vol) <- c("NeuroVol", "array") # simplified
  vol
}
```

### Basic Retrieval (Volumetric Data)

To get the default MNI T1-weighted brain template (1mm resolution):
```{r get_default_template, eval = FALSE}
# Get the default template (MNI152NLin2009cAsym T1w brain at 1mm)
mni_t1w <- get_template() 
print(mni_t1w)
```
For this vignette, we'll simulate the object:
```{r get_default_template_mock, echo=FALSE, eval=TRUE}
mni_t1w <- mock_neurovol("MNI152NLin2009cAsym_T1w")
print(mni_t1w$name) # Printing just name for brevity
```

To fetch a specific template space, suffix, and resolution:
```{r get_specific_template, eval = FALSE}
# Get MNI T1w template at 2mm resolution
mni_t1w_res2 <- get_template(
  space = "MNI152NLin2009cAsym", 
  suffix = "T1w", 
  resolution = "2" # or 2
)
print(mni_t1w_res2)
```

### Fetching Other Variants (e.g., Masks)

You can specify other TemplateFlow entities like `desc` (description) and `label`.
```{r get_mask, eval = FALSE}
# Get brain mask for MNI template
mni_brain_mask <- get_template(
  space = "MNI152NLin2009cAsym",
  desc = "brain", # 'brain' for brain masks
  suffix = "mask",
  resolution = "1"
)
print(mni_brain_mask)
```

### Getting File Paths Only

If you only need the path to the file (e.g., for use with other tools), set `path_only = TRUE`:
```{r get_path_only, eval = FALSE}
# Get just the file path, not the loaded NeuroVol object
mni_t1w_path <- get_template(
  space = "MNI152NLin2009cAsym", 
  suffix = "T1w", 
  path_only = TRUE
)
print(mni_t1w_path)
```
Example output (path will vary):
```
[1] "/Users/youruser/Library/Caches/org.R-project.R/R/neuroatlas/templateflow/tpl-MNI152NLin2009cAsym/tpl-MNI152NLin2009cAsym_res-01_T1w.nii.gz"
```

### Vectorized Retrieval

`get_template()` allows one of `space`, `variant`, `modality`, `resolution`, or `label` to be a vector, returning a named list of results.
```{r get_vectorized, eval = FALSE}
# Get both 1mm and 2mm resolution templates
t1_multi_res <- get_template(
  space = "MNI152NLin2009cAsym", 
  suffix = "T1w",
  resolution = c("1", "2") # Get both 1mm and 2mm
)
print(names(t1_multi_res))
# Access individual resolutions
print(t1_multi_res[["1"]]) # The 1mm NeuroVol
print(t1_multi_res[["2"]]) # The 2mm NeuroVol
```

### Key Parameters for `get_template()`
- `space`: Template identifier (e.g., "MNI152NLin2009cAsym").
- `variant`: High-level type (e.g., "brain", "mask", "probseg"). Used to infer `desc` and `suffix` for common cases.
- `modality`: Image type (e.g., "T1w", "T2w"). Used to infer `suffix`.
- `resolution`: Resolution (e.g., "1", "2", "01").
- `desc`: Specific TemplateFlow `desc` field.
- `suffix`: Specific TemplateFlow `suffix` field.
- `label`: E.g., for `variant="probseg"`, tissue labels like "GM", "CSF".
- `atlas`: E.g., "Schaefer2018" if fetching an atlas file.
- `...`: Other BIDS-like entities (e.g., `hemi="L"`, `den="32k"`) can be passed.

The function performs pre-flight validation checks for `space` and `resolution` against information available from TemplateFlow.

## Typed Helper Functions

For common tasks, `neuroatlas` provides convenient wrappers:

### Volumetric Data with `get_volume_template()`
```{r get_volume_helper, eval = FALSE}
# Get T1-weighted volume using the convenience function
t1_vol <- get_volume_template(
  template = "MNI152NLin2009cAsym", 
  type = "T1w", 
  resolution = "1"
)
print(t1_vol)
```
Supported `type`s include "T1w", "T2w", "bold", "probseg", "dseg".

### Surface Data with `get_surface_template()`
This helper typically returns file paths, as surfaces are often handled by specialized geometry software.
```{r get_surface_helper, eval = FALSE}
# Get fsLR left hemisphere pial surface
fslr_pial_L_path <- get_surface_template(
  template_id = "fsLR",       # e.g., "fsaverage", "fsLR"
  surface_type = "pial",      # e.g., "pial", "white", "inflated"
  hemi = "L",                 # "L" or "R"
  density = "32k"             # e.g., "32k", "164k" (for fsLR)
)
print(fslr_pial_L_path)
```
Example output (path will vary):
```
[1] "/Users/youruser/Library/Caches/org.R-project.R/R/neuroatlas/templateflow/tpl-fsLR/tpl-fsLR_den-32k_hemi-L_desc-pial_surf.gii"
```

## Caching Behavior

`neuroatlas` and TemplateFlow employ multiple layers of caching:
1.  **R-level Memoisation**: `get_template()` memoises the *file paths* it resolves. If you request the same template with identical parameters multiple times in an R session, the path is retrieved from this memory cache. If `path_only=FALSE`, `as_neurovol()` (which loads the NIfTI file) is also memoised, so the `NeuroVol` object itself is cached in memory for the session if the same path is read again.
2.  **Python Disk Cache**: The underlying Python `templateflow` library maintains a disk cache for downloaded files. `neuroatlas` configures this to use a directory like `~/.cache/neuroatlas/templateflow` (path varies by OS). Once a file is downloaded, subsequent requests (even in new R sessions) will use the disk-cached version if `TEMPLATEFLOW_HOME` remains consistent.

You can inspect and manage the `neuroatlas`-specific TemplateFlow cache:
```{r cache_management, eval = FALSE}
# Show the path to the cache directory
cache_loc <- show_templateflow_cache_path()
print(cache_loc)

# Clear the disk cache and R-level memoisation for TemplateFlow paths
# clear_templateflow_cache()                 # Asks for confirmation if interactive
# clear_templateflow_cache(confirm = FALSE)  # Clears without asking
```

## Integration with Other `neuroatlas` Functions

The `outspace` argument in functions like `get_schaefer_atlas()` can accept TemplateFlow identifiers:
```{r schaefer_tf_outspace, eval = FALSE}
# Load Schaefer atlas resampled to a TemplateFlow space
schaefer_in_mni <- get_schaefer_atlas(
  parcels = "100", 
  networks = "7",
  outspace = "MNI152NLin2009cAsym" # Resolves to default 1mm T1w brain
)

# Or be more specific with a list:
schaefer_in_mni_res2 <- get_schaefer_atlas(
  parcels = "100",
  networks = "7",
  outspace = list(space = "MNI152NLin2009cAsym", resolution = "2")
)
print(schaefer_in_mni_res2)
```
This uses an internal helper `.resolve_template_input()` to fetch the template and extract its `NeuroSpace` information.

## Summary

The `neuroatlas` package provides a comprehensive R interface to TemplateFlow:

- **Easy Setup**: Use `create_templateflow()` to initialize the interface
- **Discovery**: Find templates with `tflow_spaces()` and `tflow_files()`
- **Flexible Retrieval**: Use `get_template()` with various parameters
- **Integration**: Use TemplateFlow spaces directly in atlas functions
- **Caching**: Automatic caching at both R and Python levels

## Further Information

For more details on TemplateFlow itself, including the full range of available templates and BIDS-inspired naming conventions, please visit the [TemplateFlow website](https://www.templateflow.org/). 